### 2019-07-01 report/chapter04.Rnw
Results update
correction of tes

\chapter{Methods}
\section{Introduction and Notation}
The analysis has already been said to be restricted on clinical or health care interventions. The interventions are restricted to comparisons of two treatment arms by some measure of sanity or worsening of health. The difference in this measure between the grouops is referred to as the treatment effect. Where it is not particularly mentioned, the term treatment effect refers to any effect measure such as log risk ratio, log hazard ratio, Cohen's $d$ or Hedges $g$, Fisher's z score or Pearsons correlation coefficient. \\
Let us consider a meta-analysis with $n$ study treatment effects ($n > 1$, but typically small). A study is indexed by $i$, and it's treatment effect by  $\theta_i$. The observed treatment effect is $\hat{\theta}_i$. The pooled treatment effect of a meta-analysis will be denoted as $\theta_M$, and consequently, the observed pooled treatment effect as $\hat{\theta}_M$. Furthermore, each treatment effect is typically measured with some standard error $s_i$ and an estimate of $s_i$ is denoted as $\hat{s}_i$. The $\hat{}$ sign thus indicates if it is an estimate.\\
For continuous outcomes, let $m_t$ be the mean of the treatment group, $m_c$ the mean of the control group, and equivalently $\textrm{sd}_t$ and $\textrm{sd}_c$ the corresponding standard deviations. $n_t$ and $n_c$ are the total number of participants in the groups.
In the case of binary outcomes, let $e_t$ be the count of events in the treatment arm $e_c$ the count of events in the control group. The observed counts in a study $i$ are referred to as $e_{t,i}$ and analogously $e_{c,i}$.

\section{Effect Measures and $p$-values}
\subsection{Continuous Outcomes}
For given $(m_t, m_c), (\textrm{sd}_t, \textrm{sd}_c)$ and $(n_t, n_c)$, one can compute mean difference as well as a standardized mean difference (here: Hedges $g$) and  a standard error thereof. Note that the definition of Hedges $g$ and its standard error $s$ varies among the literature, the following applies for this report:

\begin{align}
s &= \sqrt{\frac{(n_t - 1)\textrm{sd}_t^2 + (n_c - 1)\textrm{sd}_c^2}{n_t + n_c - 2}} & g &= \frac{m_t - m_c}{s} \label{eq:hedges.g}
\end{align}

The mean difference $\theta$ and its standard error $s$ can similarly be obtained by

\begin{align}
\theta &= m_t - m_c & s &= \sqrt{\textrm{sd}_t^2/n_t + \textrm{sd}_c^2/n_t}
\end{align}

Both estimators take into account that the two groups might have unequal variances. A $p$-value to test the null hypothesis that the mean between group is equal is commonly obtained with the students $t$ test. The $t$ statistic is obtained, using $s$ and $g$ from \ref{eq:hedges.g}, by

\begin{align}
t &= g/(s\sqrt{(1/n_t)+(1/n_c)}) \nonumber
\end{align}
and the $p$-value can be obtained with the cumulative student's $t$-distribution $F$ with $n_t + n_c - 2$ degrees of freedom:

\begin{align}
p &= 2(1-F(|t|)) \nonumber
\end{align}

@@

\subsection{Binary Outcomes}
Two commonly used effect measures for binary outcome data are risk ratios and odds ratios between treatment and control group. 
The methods presented here can also be found, for example, in \cite[34]{Intro.meta}.
Let $\theta$ be the logarithm to base 10 of the odds ratio. $\theta$ and its variance $s^2$ can be obtained by

\begin{align}
\hat{\theta} &= \log(\frac{e_t*(n_c - e_c)}{e_c*(n_t - e_t)}) \nonumber \\
\hat{s}^2 &= 1/e_t + 1/(n_t - e_t) + 1/e_c + 1/(n_c - e_c) \nonumber
\end{align}

Plugging in the observed counts will give the corresponding estimates. The logarithm of the risk ratio $\theta$ and its variance $s^2$ is similarly defined as

\begin{align}
\theta &= \log(\frac{e_t/n_t}{e_c/n_c}) \label{eq:risk.ratio} \\
s^2 &= 1/e_t + 1/n_t + 1/e_c + 1/e_t \label{eq:risk.ratio.variance}
\end{align}

Using likelihood theory, one could show that the estimators are maximum likelihood estimators and that one can use the asymptotic normal distribution of the maximum likelihood estimator to calculate a $p$-value, e..g. \cite[98]{held2014}. \\
Thus, with $\Phi$ as the cumulative standard normal distribution, we get
\begin{align}
p &= 2*(1-\Phi(|\hat{\theta}/\hat{s}|) \nonumber
\end{align}
, a $p$-value for the corresponding estimates, which summarizes the evidence against $\hat{\theta}$ being zero (i.e. the true risk/odds ratio being 1). \\
Binary and continuous effect measures can be converted into each other, as described in section \ref{sec:transformation.effectsizes}.

\subsection{Survival Outcomes}
Time-to-event data with censoring has to be analyzed by special means. One frequently used method to take into account right-censoring is the Cox proportional hazards regression model \citep{Cox}. Because the method itself is not applied in this thesis, but only the resulting estimates of the parameters are used, the reader is referred to the extensive literature covering this topic (e.g. \citet{Surv}). \\
The so-called hazard ratio estimated by cox p.h. regression is the ratio of the instantaneous risk of experiencing the event between two groups. Because it is a maximum likelihood estimator, one can again use its Wald test statistic to test for equal hazards. Let $\hat{\theta}$ be an estimate of the log hazard ratio and $\hat{s}$ an estimate of the standard error of it. As before
\begin{align}
p &= 2*(1-\phi(|\hat{\theta}/\hat{s}|) \nonumber
\end{align}
will give a $p$-value for the evidence against the null hypothesis.

\section{Fixed and Random Effects Meta-Analysis} \label{sec:meta.analysis}
The fixed effects meta-analysis estimator of the pooled treatment effect is a mean of the single treatment effect estimators, weighted by their standard errors \citep{fixed.effects.rosenthal}. Let $w_i = 1/s_i^2$ be the weights, and $\theta_M$ be the pooled estimator and $s_M^2$ its variance. Then

@@

This estimator minimizes the variance between the effects. An estimate $\hat{\theta}_M$ can be obtained by plugging in the observed treatment effects and variances $\hat{\theta}_i, \hat{s}_i^2$. The underlying idea is that we assume $\theta_i \sim N(\theta_M, s_i^2)$, $\theta_M$ being the true effect, all $\theta_i$ being distributed around an equal mean. \\
The random effects model \citep{whitehead} assumes instead that 
\begin{align}
\theta_i \sim N(\mu_i, s_i^2) &&
\mu_i \sim N(\theta_M, \tau^2) \label{eq:random.effects} 
\end{align}

@@

\begin{align}
\theta_i | \mu_i \sim N(\theta_M, s_i^2 + \tau^2) \nonumber %\label{eq:random.effects.marginal}
\end{align}

$\tau^2$ is often referred to as a population variance or between-study variance, whereas $s_i^2$ can be interpreted as sampling error. The pooled treatment effect estimate $\theta_M$ of the random effects model and its variance is obtained by replacing the weights $w_i$ in equation \ref{eq:fixed.effects} with $w_i = 1/(s_i^2 + \tau^2)$. \\
The model is superior to the fixed effects model whenever the standard errors of the treatment effects alone are unlikely to fully account for the entire variability observed between studies. Note that as $\tau^2$ increases, each $\theta_i$ will eventually get equal weights, irrespective of it's sampling error $s_i^2$.\\
The estimation of $\tau^2$ has been subject to some debate in the statistical literature. Oftentimes, the method of moment estimator of \citet{tau.estimator} is used.
We use the measure of heterogeneity, $Q$, and divide by $C$ after having subtracted the degrees of freedom $n-1$:

\begin{align}
Q &= \sum_{i = 1}^n w_i(y_i - \theta_M)^2 & C &= \sum_{i = 1}^n w_i - \frac{\sum_{i = 1}^n w_i^2}{\sum_{i = 1}^n w_i}\label{eq:Q.heterogeneity} \\
\tau^2 &= \max(0, \frac{Q - (n-1)}{C}) \label{eq:Tau.definition}
\end{align}

Again, $w_i, \theta_i ...$ have to be replaced by their estimates in order to get an estimate $\hat{\tau}^2$. \\
The Paule-Mandel estimator \citep{paulemandel} is considered to have most often better properties than the method of moments estimator (e.g. \citet{tau.estimator.evaluation}).
Since we defined $w_i = 1/(s_i^2 + \tau^2)$, it als holds that

@@

\begin{align}
s^2(w_i\theta_i) &= \frac{\sum_{i = 1}^n w_i(\theta_i - \theta_M)^2}{n-1} & \frac{\sum_{i = 1}^n w_i(\theta_i - \theta_M)^2}{n-1} = 1 \label{eq:paulemandel}
\end{align}

@@

\begin{align}
I^2 &= (Q - n + 1)/Q \nonumber%\label{I2.proportion}
\end{align}

%Large values for $\hat{I}^2$ are of great importance when analyzing e.g. small study effects. \\
Importantly, all proposed methods above assume normally distributed effect sizes and proper estimates $\hat{s}$ of the true standard error. This assumptions are not met for very small sample sizes and very few event counts. Alternatively, the mantel-haenszel method for risk and odds ratios (see e.g. \citet{mantel.haenszel}) could be used in the latter case.

@@

\section{Small Study Effects Tests}
The tests that will be presented on the following pages are a common way to detect publication bias. Importantly, they are however not interpretable directly as evidence for publication bias, but this is left for the discussion chapter. The tests are also often referred to as funnel plot asymmetry tests, because of the popularity of a recent test that has been used frequently to test and adjust for publication bias, which goes under the name of trim-and-fill \citet{trimfill}. However, it is known for some time that the method has disadvantageous properties, therefore, it will not be discussed here, as well as the funnel plot (the radial plot will be used as an alternative).\\
A test for small study effects will test in some ways if the size of the estimated treatment effect of a study depends on some measure of its size. An association between study size and treatment effect size can be interpreted as an artifact of publication bias.

@@

\subsubsection{Begg and Mazumdar: Rank Correlation Test} \label{sec:Begg}
\citet{begg.ties} proposed a rank based test to test the null hypothesis of no correlation between effect size and variance.
A standardized effect size $\theta_i^\star$ can be computed as in \ref{begg.stand.eff}. $s_i^{2\star}$ is the variance of $\theta_i - \theta_M$ as defined in \ref{begg.stand.var}, $\theta_M$ being the fixed effects pooled treatment effect (\ref{eq:fixed.effects}.

\begin{align}
\theta_i^\star &= (\theta_i - \theta_M)/s_i^{2\star} \label{begg.stand.eff}  \\
s_i^{2\star} &= s_i^2 - 1/\sum_{i = 1}^n\frac{1}{s_i^2}1 \label{begg.stand.var} 
\end{align}

@@

Thus, large number of concordance between pairs will reflect in large $\hat{u}$ and small $\hat{l}$ and thus lead to a large $\hat{Z}$. The $p$-value is obtained using the standard normal distribution $\Phi$:
\begin{align}
p &= 2*(1-\Phi(|Z|)) \nonumber
\end{align}

@@

\begin{align}
\operatorname*{argmin}_{\beta_0, \beta_1}(\sum_{i = 1}^n y_i - \beta_0 - \beta_1 x_i) \label{eq:least.squares}
\end{align}

@@

\begin{align}
p &= 2(1-F(|(\beta_0 - \beta_{H0})/s_{\beta_0}|)) \nonumber
\end{align}

@@

with positive weights $w_i$ that penalize large squared residuals for some $i$ more if $w_i$ is larger compared to other $w_i$. \\
Let $\mathbf{W}$ be a $n \times n$ matrix with $\mathbf{W}_{ii} = w_i$, the weights on the diagonal and zeros on the off-diagonals. The estimates in \ref{eq:regression.parameters} and \ref{eq:regression.variances} can still be used if $\mathbf{X}$ is exchanged with $\mathbf{X}^\star =  \mathbf{W} \mathbf{X}$ and $\mathbf{y}$ with $\mathbf{y}^\star = \mathbf{W} \mathbf{y}$. \label{weighted.regression} \\
Now it is shown how linear regression can be used to test dependency of effect sizes on study sizes. The simplest application was introduced by \citet{Egger}.
Let $\theta/s$ be the dependent variable $y$ and $1/s$ the explanatory variable $x$. If plotted, this corresponds to a radial or Galbraith plot \citet{galbraith}. The linear regression equation as introduced before in \ref{eq:simple.regression} can be written in two ways:

@@

\ref{eq:radial.plot} is often provided due to the correspondence to the radial plot. However, it is equivalent to

\begin{align}
\theta &= \beta_0 + \beta_1 s + \epsilon, & \epsilon \sim N(0, w^{-1}\sigma^2) \label{eq:egger.plot}
\end{align}

with weights $w = 1/s^2$. Thus testing $\beta_0$ of \ref{eq:radial.plot} or $\beta_1$ is equivalent. The corresponding $p$-value is then used as evidence for a small study effect. Plugging in  $\theta_i/s_i, 1/s_i$ as $y_i, x_i$ into equation \ref{eq:regression.parameters} and \ref{eq:regression.variances} will give the estimates for $\hat{\beta}_0, \hat{\beta}_1,\hat{s}_{\beta_0}$ and $\hat{s}_{\beta_1}$.

@@

\begin{align}
\theta_i \sim N(\beta_0 + \beta_1 s_i, s_i^2 + \tau^2) \label{t.sharp.regression}
\end{align}

$\tau^2$ is estimated as in equation \ref{eq:Tau.definition} (method of moments). %or \ref{eq:paule.mandel}. 
The weights are set as $w_i = 1/\sqrt{s_i^2 + \tau^2})$. After adjusting for the weights as described in \ref{weighted.regression}, we can proceed analogous to Egger's test. The $p$-value for $\beta_{0} \neq 0$ reflects the evidence for a small study effect.

@@

The issue with dichotomous outcomes is that effect size and variance of effect size are correlated, which can readily be seen in \ref{eq:risk.ratio} and \ref{eq:risk.ratio.variance}. For example, a small number of event counts in one or group will inflate the variance and the effect size. Consequently, the tests above will tend to reject the null-hypothesis too often, i.e. report false positives. A number of solutions to this problem are provided in the literature.

\subsubsection{Peters Test: Weighted Linear Regression Test} \label{sec:Peter}
Instead of taking the standard error $s$ as explanatory variable $x$ as in Egger's Test, the inverse of the total sample size is used. Additionally, the variances $s_i^2$ are used as weights. Thus, the subsequent test procedure is identical to Egger's test. Peters test is a a small modification of Macaskill's test where the explanatory variable is the sample size instead of its inverse.

\subsubsection{Harbord's Test: Score based Test} \label{sec:Harbord}
A rank based alternative to Peters test for binary outcomes is Harbord's test \citep{Harbord}.
It uses a different treatment effect and variance estimate: the score $\varphi$ of the log-likelihood, evaluated at log odds ratio $\theta_0 = 0$ and its inverse Fisher information $s^2$. Formally,

\begin{align}
\varphi &= e_t - (e_t - e_c)(e_t + (n_t - e_t))/(n_t + n_c) \label{harbord.score} \\
s^2 &= \frac{(e_t + e_c)(e_t + (n_t - e_t))(e_c + (n_c - e_c))((n_t - e_t) + (n_c - e_c))}{(n_t + n_c)^2(n_t + n_c - 1)} \label{harbord.variance}
\end{align}

@@

\subsubsection{Schwarzer's Test: Rank Correlation Test} \label{sec:Schwarzer}
\citet{Schwarzer} developed a test for the correlation between $E_t - \mathbb{E}(E_t)$ and the variance of $E_t$, $E_t$ being a random variable from the non-central hypergeometric distribution, assuming a fixed log odds ratio. \\
$\mathbb{E}(E_t)$ and variance of $E_t$ are then estimated based on $e_t$. The standardized cell count deviation 
\begin{align}
(e_t - \mathbb{E}(E_t))/\sqrt(s_i^2)
\end{align}
and the inverse of $s_i^2$ is then used as before in Begg and Mazumdar's test.

@@

\begin{align}
\theta_i = \arcsin{e_t/n_t} - \arcsin{e_c/n_c} \nonumber &
s_i^2 = 1/4n_t + +/4n_c \nonumber
\end{align}

@@

\section{Excess Significance Test}
\citet{excess.significance} developped an exploratory test to detect if the proportion of significant findings is larger than expected. Note that significance is here defined as one-sided significance, with the direction of the test being the same for each study. \\
We assume that the effects are equally distributed around a true mean effect $\theta_M$, which can be estimated as for example by fixed effects Meta-Analysis \ref{eq:fixed.effects}. Let $O$ be the number of significant study results out of $n$ studies and $\alpha$ the significance threshold. Corresponding to the study effect $\theta_i$, we can specify the power $1 - \beta_i$, the probability to accepting a true result. Let $z_{\alpha,i}$ be the $1-\alpha$ quantile of a normal distribution with standard error $\hat{s}_i$. The power of study $i$ can be estimated as:

@@

with $F$ being the cumulative distribution of a normal with mean $\hat{\theta}_M$ and standard deviation $\hat{s}_i$. \\
If we assume no bias in $\theta_i, \theta_M$, the expected number of significant study results is then just

@@

and consecutively, calculating a $p$-value for the evidence against the null-hypothesis of $O = E$ with a $\chi^2$ distribution with one degree of freedom. Alternatively, one can also use a binomial proportion test, which tests if the probability of significance is the same as expected. A simple Wald test statistic $Z$ is given by:

\begin{align}
Z = \frac{O/n - E/n}{\sqrt{(O/n(1-O/n))/n}}
\end{align}

The one sided $p$-value for the alternative hypothesis $O/n > E/n$ is

\begin{align}
p &= 1 - \Phi(Z) \nonumber
\end{align}

with $\Phi$ being the cumulative distribution function of a standard normal. \citet{excess.significance} propose it to be more suitable if $n$ is small.

@@

\section{Small Study Effect and Publication Bias Adjustment}
There are different approaches to correct for small study effects and publication bias. They can mainly be distinguished by their underlying methods: regression based approaches aim to regress the effect to a study with infinite precision (i.e. very small standard error) or to a summary effect, corrected for publication bias. Selection models aim to simulate different, hypothetical selection processes and attain a approximate treatment effect by sensitivity analysis:

>>>>

\chapter{Methods} 
\section{Introduction and Notation}
The analysis has already been said to be restricted on clinical or health care interventions. The interventions are restricted to comparisons of two treatment arms by some measure of sanity or worsening of health. The difference in this measure between the grouops is referred to as the treatment effect. Where it is not particularly mentioned, the term treatment effect refers to any effect measure such as log risk ratio, log hazard ratio, Cohen's $d$ or Hedges $g$, Fisher's z score or Pearsons correlation coefficient. \\
Let us consider a meta-analysis with $n$ study treatment effects ($n > 1$, but typically small). A study is indexed by $i$, and it's treatment effect by  $\theta_i$. The observed treatment effect is $\hat{\theta}_i$. The pooled treatment effect of a meta-analysis will be denoted as $\theta_M$, and consequently, the observed pooled treatment effect as $\hat{\theta}_M$. Furthermore, each treatment effect is typically measured with some standard error $\se_i$ and an estimate of $\se_i$ is denoted as $\hat{\se}_i$. The $\hat{}$ sign thus indicates if it is an estimate.\\
For continuous outcomes, let $m_t$ be the mean of the treatment group, $m_c$ the mean of the control group, and equivalently $\textrm{sd}_t$ and $\textrm{sd}_c$ the corresponding standard deviations. $n_t$ and $n_c$ are the total number of participants in the groups.
In the case of binary outcomes, let $e_t$ be the count of events in the treatment arm $e_c$ the count of events in the control group. The observed counts in a study $i$ are referred to as $e_{t,i}$ and analogously $e_{c,i}$.

\section{Effect Measures and $p$-values}
\subsection{Continuous Outcomes}
For given $(m_t, m_c), (\textrm{sd}_t, \textrm{sd}_c)$ and $(n_t, n_c)$, one can compute mean difference as well as a standardized mean difference (here: Hedges $g$) and  a standard error thereof. Note that the definition of Hedges $g$ and its standard error $\se$ varies among the literature, the following applies for this report:

\begin{align}
\se &= \sqrt{\frac{(n_t - 1)\textrm{sd}_t^2 + (n_c - 1)\textrm{sd}_c^2}{n_t + n_c - 2}} & g &= \frac{m_t - m_c}{\se} \label{eq:hedges.g}
\end{align}

The mean difference $\theta$ and its standard error $\se$ can similarly be obtained by

\begin{align}
\theta &= m_t - m_c & \se &= \sqrt{\textrm{sd}_t^2/n_t + \textrm{sd}_c^2/n_t}
\end{align}

Both estimators take into account that the two groups might have unequal variances. A $p$-value to test the null hypothesis that the mean between group is equal is commonly obtained with the Students $t$ test. The $t$ statistic is obtained, using $s$ and $g$ from \ref{eq:hedges.g}, by

\begin{align}
t &= g/(s\sqrt{(1/n_t)+(1/n_c)}) \nonumber
\end{align}
and the $p$-value can be obtained with the cumulative Student's $t$-distribution $F$ with $n_t + n_c - 2$ degrees of freedom:

\begin{align}
p &= 2(1-F(\abs{t})) \nonumber
\end{align}

@@

\subsection{Binary Outcomes}
Two commonly used effect measures for binary outcome data are risk ratios and odds ratios between treatment and control group. 
The methods presented here can also be found, for example, in \cite[34]{Intro.meta}.
Let $\theta$ be the natural logarithm of the odds ratio. $\theta$ and its variance $\se^2$ can be obtained by

\begin{align}
\theta &= \log(\frac{e_t\cdot(n_c - e_c)}{e_c\cdot(n_t - e_t)}) \nonumber \\
\se^2 &= 1/e_t + 1/(n_t - e_t) + 1/e_c + 1/(n_c - e_c) \nonumber
\end{align}

Plugging in the observed counts will give the corresponding estimates. The logarithm of the risk ratio $\theta$ and its variance $\se^2$ is similarly defined as

\begin{align}
\theta &= \log(\frac{e_t/n_t}{e_c/n_c}) \label{eq:risk.ratio} \\
\se^2&= 1/e_t + 1/n_t + 1/e_c + 1/e_t \label{eq:risk.ratio.variance}
\end{align}

Using likelihood theory, one could show that the estimators are maximum likelihood estimators and that one can use the asymptotic normal distribution of the maximum likelihood estimator to calculate a $p$-value, \eg \cite[98]{held2014}. \\
Thus, with $\Phi$ as the cumulative standard normal distribution, we get
\begin{align}
p &= 2\cdot(1-\Phi(\abs{\hat{\theta}/\hat{\se}}), \nonumber
\end{align}
a $p$-value for the corresponding estimate, which summarizes the evidence against $\hat{\theta}$ being zero (\ie the true risk/odds ratio being 1). \\
Binary and continuous effect measures can be converted into each other, as described in Section \ref{sec:transformation.effectsizes}.

\subsection{Survival Outcomes}
Time-to-event data with censoring has to be analyzed by special means. One frequently used method to take into account right-censoring is the Cox proportional hazards regression model \citep{Cox}. Because the method itself is not applied in this thesis, but only the resulting estimates of the parameters are used, the reader is referred to the extensive literature covering this topic (\eg \citet{Surv}). \\
The so-called hazard ratio estimated by Cox regression is the ratio of the instantaneous risk of experiencing the event between two groups. Because it is a maximum likelihood estimator, one can again use its Wald test statistic to test for equal hazards. Let $\hat{\theta}$ be an estimate of the log hazard ratio and $\hat{\se}$ an estimate of the standard error of it. As before
\begin{align}
p &= 2\cdot(1-\phi(\abs{\hat{\theta}/\hat{\se}(\hat{\theta})}) \nonumber
\end{align}
will give a $p$-value for the evidence against the null hypothesis.

\section{Fixed and Random Effects Meta-Analysis} \label{sec:meta.analysis}
The fixed effects meta-analysis estimator of the pooled treatment effect is a mean of the single treatment effect estimators, weighted by their standard errors \citep{fixed.effects.rosenthal}. Let $w_i = 1/\se_i^2$ be the weights, and $\theta_M$ be the pooled estimator and $s_M^2$ its variance. Then

@@

This estimator minimizes the variance between the effects. An estimate $\hat{\theta}_M$ can be obtained by plugging in the observed treatment effects and variances $\hat{\theta}_i$ and $\hat{\se}_i^2$. The underlying idea is that we assume $\theta_i \sim N(\theta_M, \se_i^2)$, $\theta_M$ being the true effect, all $\theta_i$ being distributed around an equal mean. \\
The random effects model \citep{whitehead} assumes instead that 
\begin{align}
\theta_i \sim N(\mu_i, \se_i^2) &&
\mu_i \sim N(\theta_M, \tau^2) \label{eq:random.effects} 
\end{align}

@@

\begin{align}
\theta_i \given \mu_i \sim N(\theta_M, \se_i^2 + \tau^2) \nonumber %\label{eq:random.effects.marginal}
\end{align}

$\tau^2$ is often referred to as a population variance or between-study variance, whereas $\se_i^2$ can be interpreted as sampling error. The pooled treatment effect estimate $\theta_M$ of the random effects model and its variance is obtained by replacing the weights $w_i$ in equation \ref{eq:fixed.effects} with $w_i = 1/(\se_i^2 + \tau^2)$. \\
The model is superior to the fixed effects model whenever the standard errors of the treatment effects alone are unlikely to fully account for the entire variability observed between studies. Note that as $\tau^2$ increases, each $\theta_i$ will eventually get equal weights, irrespective of it's sampling error $\se_i^2$.\\
The estimation of $\tau^2$ has been subject to some debate in the statistical literature. Oftentimes, the method of moment estimator of \citet{tau.estimator} is used.
We use the measure of heterogeneity, $Q$, and divide by $C$ after having subtracted the degrees of freedom $n-1$:

\begin{align}
Q &= \sum_{i = 1}^n w_i(y_i - \theta_M)^2 & C &= \sum_{i = 1}^n w_i - \frac{\sum_{i = 1}^n w_i^2}{\sum_{i = 1}^n w_i}\label{eq:Q.heterogeneity} \\
\tau^2 &= \max\big(0, \frac{Q - (n-1)}{C}\big) \label{eq:Tau.definition}
\end{align}

Again, $w_i, \theta_i ...$ have to be replaced by their estimates in order to get an estimate $\hat{\tau}^2$. \\
The Paule-Mandel estimator \citep{paulemandel} is considered to have most often better properties than the method of moments estimator (\eg \citet{tau.estimator.evaluation}).
Since we defined $w_i = 1/(\se_i^2 + \tau^2)$, it als holds that

@@

\begin{align}
\se^2(w_i\theta_i) &= \frac{\sum_{i = 1}^n w_i(\theta_i - \theta_M)^2}{n-1} & \frac{\sum_{i = 1}^n w_i(\theta_i - \theta_M)^2}{n-1} = 1 \label{eq:paulemandel}
\end{align}

@@

\begin{align}
I^2 &= \max\big(0, 1 - \frac{n-1}{Q}\big) \nonumber%\label{I2.proportion}
\end{align}

%Large values for $\hat{I}^2$ are of great importance when analyzing \eg small study effects. \\
Importantly, all proposed methods above assume normally distributed effect sizes and proper estimates $\hat{\se}$ of the true standard error. This assumptions are not met for very small sample sizes and very few event counts. Alternatively, the Mantel-Haenszel method for risk and odds ratios (see \eg \citet{mantel.haenszel}) could be used in the latter case.

@@

\section{Small Study Effects Tests}
The tests that will be presented on the following pages are a common way to detect publication bias. Importantly, they are however not interpretable directly as evidence for publication bias, but this is left for the discussion chapter. The tests are also often referred to as funnel plot asymmetry tests, because of the popularity of a recent test that has been used frequently to test and adjust for publication bias, which goes under the name of trim-and-fill \citep{trimfill}. However, it is known for some time that the method has disadvantageous properties, therefore, it will not be discussed here, as well as the funnel plot (the radial plot will be used as an alternative).\\
A test for small study effects will test in some ways if the size of the estimated treatment effect of a study depends on some measure of its size. An association between study size and treatment effect size can be interpreted as an artifact of publication bias.

@@

\subsubsection{Begg and Mazumdar: Rank Correlation Test} \label{sec:Begg}
\citet{begg.ties} proposed a rank based test to test the null hypothesis of no correlation between effect size and variance.
A standardized effect size $\theta_i^\star$ can be computed as in \ref{begg.stand.eff}. $\se_i^{2\star}$ is the variance of $\theta_i - \theta_M$ as defined in \ref{begg.stand.var}, $\theta_M$ being the fixed effects pooled treatment effect (\ref{eq:fixed.effects}.

\begin{align}
\theta_i^\star &= (\theta_i - \theta_M)/\se_i^{2\star} \label{begg.stand.eff}  \\
\se_i^{2\star} &= \se_i^2 - 1/\sum_{i = 1}^n\frac{1}{\se_i^2}1 \label{begg.stand.var} 
\end{align}

@@

Thus, large number of concordance between pairs will reflect in large $\hat{u}$ and small $\hat{l}$ and thus lead to a large $\hat{Z}$. The $p$-value is obtained using the standard normal distribution $\Phi$:
\begin{align}
p &= 2\cdot(1-\Phi(\abs{Z})) \nonumber
\end{align}

@@

\begin{align}
\operatorname*{argmin}_{\beta_0, \beta_1}\big(\sum_{i = 1}^n y_i - \beta_0 - \beta_1 x_i\big) \label{eq:least.squares}
\end{align}

@@

\begin{align}
p &= 2(1-F(\abs{(\beta_0 - \beta_{H0})/s_{\beta_0}})) \nonumber
\end{align}

@@

with positive weights $w_i$ that penalize large squared residuals for some $i$ more if $w_i$ is larger compared to other $w_i$. \\
Let $\mathbf{W}$ be a $n \times n$ matrix with $\mathbf{W}_{ii} = w_i$, the weights on the diagonal and zeros on the off-diagonals. The estimates in \ref{eq:regression.parameters} and \ref{eq:regression.variances} can still be used if $\mathbf{X}$ is exchanged with $\mathbf{X}^\star =  \mathbf{W} \mathbf{X}$ and $\mathbf{y}$ with $\mathbf{y}^\star = \mathbf{W} \mathbf{y}$. \label{weighted.regression} \\
Now it is shown how linear regression can be used to test dependency of effect sizes on study sizes. The simplest application was introduced by \citet{Egger}.
Let $\theta/s$ be the dependent variable $y$ and $1/s$ the explanatory variable $x$. If plotted, this corresponds to a radial or Galbraith plot \citep{galbraith}. The linear regression equation as introduced before in \ref{eq:simple.regression} can be written in two ways:

@@

Equation \eqref{eq:radial.plot} is often provided due to the correspondence to the radial plot. However, it is equivalent to

\begin{align}
\theta &= \beta_0 + \beta_1 \se + \epsilon, & \epsilon \sim N(0, w^{-1}\sigma^2) \label{eq:egger.plot}
\end{align}

with weights $w = 1/\se^2$. Thus testing $\beta_0$ of \ref{eq:radial.plot} or $\beta_1$ is equivalent. The corresponding $p$-value is then used as evidence for a small study effect. Plugging in  $\theta_i/\se_i, 1/\se_i$ as $y_i, x_i$ into equation \ref{eq:regression.parameters} and \ref{eq:regression.variances} will give the estimates for $\hat{\beta}_0, \hat{\beta}_1,\hat{\se}_{\beta_0}$ and $\hat{\se}_{\beta_1}$.

@@

\begin{align}
\theta_i \sim N(\beta_0 + \beta_1 \se_i, \se_i^2 + \tau^2) \label{t.sharp.regression}
\end{align}

$\tau^2$ is estimated as in equation \ref{eq:Tau.definition} (method of moments). %or \ref{eq:paule.mandel}. 
The weights are set as $w_i = 1/\sqrt{\se_i^2 + \tau^2})$. After adjusting for the weights as described in \ref{weighted.regression}, we can proceed analogous to Egger's test. The $p$-value for $\beta_{0} \neq 0$ reflects the evidence for a small study effect.

@@

The issue with dichotomous outcomes is that effect size and variance of effect size are correlated, which can readily be seen in \ref{eq:risk.ratio} and \ref{eq:risk.ratio.variance}. For example, a small number of event counts in one or group will inflate the variance and the effect size. Consequently, the tests above will tend to reject the null-hypothesis too often, \ie report false positives. A number of solutions to this problem are provided in the literature.

\subsubsection{Peters Test: Weighted Linear Regression Test} \label{sec:Peter}
Instead of taking the standard error $s$ as explanatory variable $x$ as in Egger's Test, the inverse of the total sample size is used. Additionally, the variances $\se_i^2$ are used as weights. Thus, the subsequent test procedure is identical to Egger's test. Peters test is a a small modification of Macaskill's test where the explanatory variable is the sample size instead of its inverse.

\subsubsection{Harbord's Test: Score based Test} \label{sec:Harbord}
A rank based alternative to Peters test for binary outcomes is Harbord's test \citep{Harbord}.
It uses a different treatment effect and variance estimate: the score $\varphi$ of the log-likelihood, evaluated at log odds ratio $\theta_0 = 0$ and its inverse Fisher information $\se^2$. Formally,

\begin{align}
\varphi &= e_t - (e_t - e_c)(e_t + (n_t - e_t))/(n_t + n_c) \label{harbord.score} \\
 \se^2&= \frac{(e_t + e_c)(e_t + (n_t - e_t))(e_c + (n_c - e_c))((n_t - e_t) + (n_c - e_c))}{(n_t + n_c)^2(n_t + n_c - 1)} \label{harbord.variance}
\end{align}

@@

\subsubsection{Schwarzer's Test: Rank Correlation Test} \label{sec:Schwarzer}
\citet{Schwarzer} developed a test for the correlation between $E_t - \mathbb{E}(E_t)$ and the variance of $E_t$, $E_t$ being a random variable from the non-central hypergeometric distribution, assuming a fixed log odds ratio. \\
$\mathbb{E}(E_t)$ and variance of $E_t$ are then estimated based on $e_t$. The standardized cell count deviation 
\begin{align}
(e_t - \mathbb{E}(E_t))/\sqrt(\se_i^2)
\end{align}
and the inverse of $\se_i^2$ is then used as before in Begg and Mazumdar's test.

@@

\begin{align}
\theta_i &= \arcsin{e_t/n_t} - \arcsin{e_c/n_c} \nonumber &
\se_i^2 &= 1/4n_t + /4n_c \nonumber
\end{align}

@@

\subsection{Excess Significance Test} \label{sec:excess.significance}
\citet{excess.significance} developped an exploratory test to detect if the proportion of significant findings is larger than expected. Note that significance is here defined as one-sided significance, with the direction of the test being the same for each study. \\
We assume that the effects are equally distributed around a true mean effect $\theta_M$, which can be estimated by fixed effects Meta-Analysis \ref{eq:fixed.effects}. Let $O$ be the number of significant study results out of $n$ studies and $\alpha$ the significance threshold. Corresponding to the study effect $\theta_i$, we can specify the power $1 - \beta_i$, the probability to accepting a true result. Let $z_{\alpha,i}$ be the $1-\alpha$ quantile of a normal distribution with standard error $\hat{\se}_i$. The power of study $i$ can be estimated as:

@@

with $F$ being the cumulative distribution of a normal with mean $\hat{\theta}_M$ and standard deviation $\hat{\se}_i$. \\
If we assume no bias in $\theta_i$ and $\theta_M$, the expected number of significant study results is then just

@@

and consecutively, calculating a $p$-value for the evidence against the null-hypothesis of $O = E$ with a $\chi^2$ distribution with one degree of freedom. Alternatively, one can also use a binomial test, which is encouraged when $n$ and $O$ is small. We will get a one sided $p$-value for excess significance, $\P(X \geq O)$, by

\begin{align}
p &= \sum_{i = O}^n(n \choose i p^i (1-p)^{n - i})
\end{align}

with $p = E/n$ and $X$ being a binomial random variable with probability $p$.

@@

\section{Small Study Effect and Publication Bias Adjustment}
There are different approaches to correct for small study effects and publication bias. They can mainly be distinguished by their underlying methods: regression based approaches aim to regress the effect to a study with infinite precision (\ie very small standard error) or to a summary effect, corrected for publication bias. Selection models aim to simulate different, hypothetical selection processes and attain a approximate treatment effect by sensitivity analysis:
